### 2015-06-17 chapters/ipa_meets_unicode.tex
update section missing decomposition

One important distinction to acknowledge is the difference between multigraphs
and ligatures. Multigraphs are groups of characters (in the context of IPA e.g.
<tʃ> or <ou>) while ligatures are single characters (e.g. <ʧ> \textsc{latin
small letter tesh digraph} at \uni{02A7}). Ligatures arose in the context of
printing easier-to-read texts, and are included in the Unicode Standard for
reasons of legacy encoding. However, their usage is discouraged by the Unicode
core specification. Specifically related to IPA, various phonetic combinations
of characters (typically affricates) are available as single code-points in the
Unicode Standard, but are designated as \textsc{ligatures} or \textsc{digraphs}
(confusingly both names appear interchangeably). Such glyphs might be used by
software to produce a pleasing display, but they should not be hard-coded into
the text itself. In the context of IPA, characters like the following ligatures
should thus \emph{not} be used. Instead a combination of two characters is
preferred:

@@

Although many combinations of base character with diacritic are treated as 
canonical equivalent with precomposed characters, there are a few combinations 
in IPA that allow for multiple, apparently identical, encodings that are not 
canonical equivalent. The following elements should not be treated as diacritics 
when encoding IPA in Unicode:
\begin{itemize}
  \item[] <\ {\large  ̡}\ > \textsc{combining palatalized hook below} at \uni{0321}
  \item[] <\ \ {\large  ̢}> \textsc{combining retroflex hook below} at \uni{0322}
  \item[] <\ \ {\large  ̵}> \textsc{combining short stroke overlay} at \uni{0335}
  \item[] <\ \ {\large  ̷}> \textsc{combining short solidus overlay} at \uni{0337}
  \item[] <\ \ {\large  ̴}> \textsc{combining tilde overlay} at \uni{0334}
  \item[] <{\large ˞}> \textsc{modifier letter rhotic hook} at \uni{02DE}
\end{itemize}

There turn out to be a lot of characters in the IPA that could be conceived as 
using any of these elements, like <ɲ>, <ɳ>, <ɨ>, <ø>, <ɫ> or <ɚ>. However, all 
such characters exist as well as ``precomposed'' combination in Unicode, and these 
precomposed characters should preferably be used. When combinations of a base 
character with diacritic are used, then these combinations are not canonical 
equivalent to the precomposed combinations. This means that any search will not 
find both at the same time.

Reversely, <ç> \textsc{latin small letter c with cedilla} at \uni{00E7} will be
decomposed into <c> \textsc{latin small letter c} at \uni{0063} with 
<\ \ {\large  ̧}> \textsc{combining cedilla} at \uni{0327}, 
also if such a decomposition is not intended, because it is meaningless in 
IPA.\@

>>>>

One important distinction to acknowledge is the difference between multigraphs
and ligatures. Multigraphs are groups of characters (in the context of IPA e.g.
<tʃ> or <ou>) while ligatures are single characters (e.g. <ʧ> \textsc{latin
small letter tesh digraph} at \uni{02A7}). Ligatures arose in the context of
printing easier-to-read texts, and are included in the Unicode Standard for
reasons of legacy encoding. However, their usage is discouraged by the Unicode
core specification. Specifically related to IPA, various phonetic combinations
of characters (typically affricates) are available as single code-points in the
Unicode Standard, but are designated \textsc{digraphs}. Such glyphs might be used by
software to produce a pleasing display, but they should not be hard-coded into
the text itself. In the context of IPA, characters like the following ligatures
should thus \emph{not} be used. Instead a combination of two characters is
preferred:

@@

Although many combinations of base character with diacritic are treated as
canonical equivalent with precomposed characters, there are a few combinations
in IPA that allow for multiple, apparently identical, encodings that are not
canonical equivalent (see Section~\ref{pitfall-canonical-equivalence}),. The
following elements should not be treated as diacritics when encoding IPA in
Unicode: 
\begin{itemize}
  \item[] <\ {\large  ̡}\ > \textsc{combining palatalized hook below} at \uni{0321}
  \item[] <\ \ {\large  ̢}> \textsc{combining retroflex hook below} at \uni{0322}
  \item[] <\ \ {\large  ̵}> \textsc{combining short stroke overlay} at \uni{0335}
  \item[] <\ \ {\large  ̷}> \textsc{combining short solidus overlay} at \uni{0337}
  \item[] <\ \ {\large  ̴}> \textsc{combining tilde overlay} at \uni{0334}
\end{itemize}

There turn out to be a lot of characters in the IPA that could be conceived as 
using any of these elements, like <ɲ>, <ɳ>, <ɨ>, <ø> or <ɫ>. However, all 
such characters exist as well as precomposed combination in Unicode, and these 
precomposed characters should preferably be used. When combinations of a base 
character with diacritic are used, then these combinations are not canonical 
equivalent to the precomposed combinations. This means that any search will not 
find both at the same time.

A similar problem arises with the rhotic hook. There are two precomposed
characters in Unicode with a rhotic hook, which are not canonical equivalent 
with a combination of the vowel with a separately encoded hook:
\begin{itemize}
  \item[] <ɚ> \textsc{latin small letter schwa with hook} at \uni{025A}
  \item[] <ɝ> \textsc{latin small letter reversed open e with hook}at \uni{025D}
\end{itemize}

All other combinations of vowels with rhotic hooks will have to be made by using
<{\large ˞}> \textsc{modifier letter rhotic hook} at \uni{02DE}, because there
is not complete set of precomposed characters with rhotic hooks. For that reason
we propose to not use the two precomposed characters with hooks mentioned above,
but always use the separate rhotic hook at \uni{02DE} in IPA.\@

Reversely, note that the <ç> \textsc{latin small letter c with cedilla} at
\uni{00E7} will be decomposed into <c> with <\ \ {\large ̧}> \textsc{combining
cedilla} at \uni{0327} by Unicode canonical decomposition, also if such a
decomposition is not intended in the IPA.\@
